
%%%%%%%%%%%%% Acknowledgements

\chapter*{Acknowledgements}

Writing a thesis in 2020 has been an incredibly challenging task. Academia can feel, at times, rather lonesome. The difficulties that emerged this year could have made the experience much more arduous. However, the completion of this body of work is a testament to the network of support I have found myself privileged to have. There were times where the motivation to do the work simply could not be found. However, knowing that I had people around me who believed in me and the work I was doing helped me get to the finish line.


To my father and mother, Alexander and Jocelyn. I hope I’ve made you proud. Being the last child to go to university and then continue onto post-graduation studies hopefully makes up for that. You have both supported me in all of my endeavours, both foolish and worthwhile. For this, I will always be thankful. Maybe one day you will finally have a doctor in the family*. To my siblings Joyce and Jon. You both are an inspiration to me because of the respective paths you have chosen.


To my fellow MRes cohort. I’m fortunate to have gone through this research journey with you all during these strange times. Sure, we’ve missed out on the usual journey one gets to experience when completing higher degree research, but I feel that the support we have given each other has easily outweighed the burdens placed upon us. I also would like to thank Dr. Michelle Jamieson and everyone in the writing group. You have all helped me craft my skill in writing, structuring and editing my rambling thoughts into somewhat coherent text. To my German lecturers and tutors, Dr. Eva Schmidt, Ms. Susana Catalina Prat and Dr. Brangwen Stone. Thank you all for nurturing my fascination with the German language.


I would like to express the sincerest sense of gratitude to my supervisors Assoc. Prof. Ulrike Garde and Dr. Maryam Khalid. Your constant encouragement made this thesis possible. I was worried when I chose my research topic and the methodology as it was a harder path to follow. However, neither of you doubted my ability to see it through. I hope I’ve made you both new fans of the world game in the process.


Finally, I would like to thank Nikki, my better half and the mother of our four-legged prince, Chuck. You have been the rock of my life. You inspired me to go to university all those years ago. You inspired me to continue through to post-graduation studies. You continue to inspire me. Thank you for guiding me through this entire process, reading my work and challenging me. I know I’m not the greatest at showing appreciation, but please know that I am grateful to you. My life would be poorer without you and your support. Thank you.

\vfill
*Yes, you can count Nikki as a doctor in the family if you want.
